---
title: 'TCGA Workflow: Analyze cancer genomics and epigenomics data using Bioconductor
  packages'
author: Tiago C. Silva, Antonio Colaprico, Catharina Olsen, Fulvio D’Angelo, Gianluca
  Bontempi Michele Ceccarelli , and Houtan Noushmehr
date: "`r Sys.Date()`"
bibliography: bibliography.bib

vignette: >
  %\VignetteIndexEntry{'TCGA Workflow: Analyze cancer genomics and epigenomics data using Bioconductor packages'}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
  %\VignetteDepends{AnnotationHub}
---

```{r setup, include=FALSE, echo=F, warning= F, message=F}
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE, 
                      error = FALSE, 
                      tidy = FALSE,
                      fig.align="center", 
                      dpi = 300, 
                      cache = FALSE,
                      progress=FALSE, 
                      quite = TRUE)
```

```{r, echo=FALSE, results="hide", warning=FALSE,message=FALSE}
suppressPackageStartupMessages({
  if("TCGAWorkflow" %in% installed.packages()[,1]) {
    library(TCGAWorkflow)
  } else {
    devtools::load_all(".")
  }
})
```

```{R, eval=TRUE, message=FALSE, warning=FALSE, include=FALSE}
library(TCGAWorkflowData)
library(DT)
library(TCGAbiolinks)
```


# Methods

![ROC with corresponding AUC for inferred GBM networks compared to
    BioGrid interactions\label{fig::roc_GBM}](roc_biogrid_GBM.png)

## Epigenetic analysis

The DNA methylation is an important component in numerous cellular
processes, such as embryonic development, genomic imprinting,
X-chromosome inactivation, and preservation of chromosome stability
[@phillips2008role].

In mammals DNA methylation is found sparsely but globally, distributed
in definite CpG sequences throughout the entire genome; however, there
is an exception. CpG islands (CGIs) which are short interspersed DNA
sequences that are enriched for GC. These islands are normally found in
sites of transcription initiation and their methylation can lead to gene
silencing [@deaton2011cpg].

Thus, the investigation of the DNA methylation is crucial to
understanding regulatory gene networks in cancer as the DNA methylation
represses transcription [@robertson2005dna]. Therefore, the DMR
(Differentially Methylation Region) detection can help us investigate
regulatory gene networks.

This section describes the analysis of DNA methylation using the
Bioconductor package
[TCGAbiolinks](http://bioconductor.org/packages/TCGAbiolinks/)
[@TCGAbiolinks]. For this analysis, and due to the time required to
perform it, we selected only 10 LGG samples and 10 GBM samples that have
both DNA methylation data from Infinium HumanMethylation450 and gene
expression from Illumina HiSeq 2000 RNA Sequencing Version 2 analysis
(lines 1-56 of the listing below describes how to make the
data acquisition). We started by checking the mean DNA methylation of
different groups of samples, then performed a DMR in which we search for
regions of possible biological significance, (e.g., regions that are
methylated in one group and unmethylated in the other). After finding
these regions, they can be visualized using heatmaps.

### Visualizing the mean DNA methylation of each patient

It should be highlighted that some pre-processing of the DNA methylation
data was done. The DNA methylation data from the 450k platform has three
types of probes cg (CpG loci) , ch (non-CpG loci) and rs (SNP assay).
The last type of probe can be used for sample identification and
tracking and should be excluded for differential methylation analysis
according to the [ilumina
manual](https://support.illumina.com/content/dam/illumina-support/documents/myillumina/90666eaa-0c66-48b4-8199-3be99b2b3ef9/genomestudio_methylation_v1.8_user_guide_11319130_b.pdf).
Therefore, the rs probes were removed (see listing below lines
68). Also, probes in chromosomes X, Y were removed to eliminate
potential artifacts originating from the presence of a different
proportion of males and females [@marabita2013evaluation]. The last
pre-processing steps were to remove probes with at least one NA value
(see listing below lines 65).

After this pre-processing step and using the function
*TCGAvisualize\_meanMethylation* function, we can look at the mean DNA
methylation of each patient in each group. It receives as argument a
*SummarizedExperiment* object with the DNA methylation data, and the
arguments *groupCol* and *subgroupCol* which should be two columns from
the sample information matrix of the *SummarizedExperiment* object
(accessed by the *colData* function) (see listing below lines
70-74).

```{R, eval=FALSE, include=TRUE}
#----------------------------
# Obtaining DNA methylation
#----------------------------
# Samples
lgg.samples <- matched_met_exp("TCGA-LGG", n = 10)
gbm.samples <- matched_met_exp("TCGA-GBM", n = 10)
samples <- c(lgg.samples,gbm.samples)

#-----------------------------------
# 1 - Methylation
# ----------------------------------
# For methylation it is quicker in this case to download the tar.gz file
# and get the samples we want instead of downloading files by files
query <- GDCquery(project = c("TCGA-LGG","TCGA-GBM"),
                  data.category = "DNA methylation",
                  platform = "Illumina Human Methylation 450",
                  legacy = TRUE, 
                  barcode = samples)
GDCdownload(query)
met <- GDCprepare(query, save = FALSE)

# We will use only chr9 to make the example faster
met <- subset(met,subset = as.character(seqnames(met)) %in% c("chr9"))
# This data is avaliable in the package
save("met.20.samples.GBM.LGG.chr9.rda")
```

```{r  echo=TRUE, message=FALSE,warning=FALSE}
library(TCGAWorkflowData)
data("met.20.samples.GBM.LGG.chr9")
#----------------------------
# Mean methylation
#----------------------------
# Plot a barplot for the groups in the disease column in the
# summarizedExperiment object

# remove probes with NA (similar to na.omit)
met <- subset(met,subset = (rowSums(is.na(assay(met))) == 0))

TCGAvisualize_meanMethylation(met,
                              groupCol = "disease_type",
                              group.legend  = "Groups",
                              filename = "mean_lgg_gbm.png",
                              print.pvalue = TRUE)
```

The figure below illustrates a mean DNA methylation plot for each
sample in the GBM group (140 samples) and a mean DNA methylation for
each sample in the LGG group. Genome-wide view of the data highlights a
difference between the groups of tumors.

![Boxplot of mean DNA methylation of each sample (black
dots)](mean_lgg_gbm.png)
